---
title: "Trait Selection"
author: Anna L Tyler
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
  html_document:
    code_folding: hide
    collapsed: no
    toc: yes
    toc_float: yes
---

## Introduction
This workflow selects traits for use in high-dimensional mediation.
The final product is a trait matrix that is used in downstream analyses.

```{r load_code}
is.interactive = FALSE
#is.interactive = TRUE
library("here")

all.fun <- list.files(here("Code"), full.names = TRUE, pattern = ".R")
for(i in 1:length(all.fun)){source(all.fun[i])}
```

```{r load_libraries, message = FALSE, warning = FALSE, error = FALSE}
all.packages <- c("pheatmap")
load_libraries(all.packages, personal.library = TRUE)
```

## Data

Data were generated through a collaboration between the Attie
lab at the University of Wisconsin, and The Jackson Laboratory
Detailed descriptions of mouse handling, phenotype gathering,
and initial data processing can be found elsewhere. 

Here we focus exclusively on the clinical phenotoype data.

## Read phenotypes and adjust

Adjust phenotype data for DO generation, sex, and diet days.

```{r pheno}
pheno.data <- readRDS(here("Data", "DO", "Clinical_Phenotypes_V11.RDS"))
#write out annotations for phenotypes
write.table(pheno.data$annot.phenotype, here("Data", "DO", "Phenotype_Annotations.txt"), 
  sep = "\t", quote = FALSE, row.names = FALSE)


num.pheno <- pheno.data$data
#fix the spelling of adiposity
colnames(num.pheno)[which(colnames(num.pheno) == "adipocity")] <- "adiposity"
covar <- as.matrix(pheno.data$annot.samples[,c("Sex", "Wave", "Generation")])
num.covar <- dummy_covar(covar)
rownames(num.covar) <- rownames(num.pheno)
saveRDS(num.covar, here("Data", "DO", "Clinical_Phenotype_Covariates.RDS"))
adj.pheno <- adjust(num.pheno, num.covar) #remove sex, wave, and generation from the phenotypes
```

## Body Weight

Investigate body weight patterns.
The following boxplot shows two wicked outliers that 
are definitely mistakes. I am removing the point in 
week 11 that is 200 grams, and the other in week 17 
that is 5 grams. 

```{r bw}
weight.idx <- grep("weight", colnames(adj.pheno))
#remove organ weights
organ.idx <- sapply(c("heart", "liver", "fat"), function(x) grep(x, colnames(adj.pheno)))
bw.idx <- setdiff(weight.idx, organ.idx)
bw <- adj.pheno[,bw.idx]
ordered.bw  <- cbind(bw[,-1], bw[,1,drop=FALSE])
boxplot(ordered.bw, las = 2)
```

The following boxplot shows the corrected weights

```{r remove_obvious_errors}
outlier1 <- which(ordered.bw > 100)
ordered.bw[outlier1] <- NA

outlier2 <- which(ordered.bw < 6)
ordered.bw[outlier2] <- NA
boxplot(ordered.bw, las = 2)
```

There is really one axis of variation in the weight
matrix that corresponds very highly with the final 
weight. The following barplot shows the variance
explained by the PCs of the weight matrix. 

```{r weight_decomp, warning = FALSE}
bw.decomp <- plot.decomp(ordered.bw, plot.results = FALSE)
barplot_with_num(round(bw.decomp$var.exp, 2), main = "Variance Explained",
  text.srt = 45)
```

The following plot shows the high concordance between the first 
PC of the weight matrix and the final weight. Thus, I think we
can safely use final weight as the only weight measurement
without losing any important information.

```{r weight_pc}
avg.weight <- apply(ordered.bw, 1, function(x) mean(x, na.rm = TRUE))
plot.with.model(bw.decomp$u[,1], ordered.bw[bw.decomp$rows.used,"weight_final"],
  xlab = "PC1 of weight matrix", ylab = "Final Weight")

#plot.with.model(bw.decomp$u[,1], avg.weight[bw.decomp$rows.used],
#  xlab = "PC1 of weight matrix", ylab = "Average Weight")
```

```{r final_trait_mat}
#initialize a trait matrix that contains all the traits we want to examine.
final.trait.mat <- adj.pheno[,"weight_final",drop=FALSE]
```

## Food consumption

We also looked at the range of food consumption.
The following boxplot shows that there is a clear
error in the food consumption measurements. We
removed the measurement that was greater than 40.

```{r food}
food.idx <- grep("food", colnames(num.pheno))
food.mat <- num.pheno[,food.idx]
boxplot(food.mat, las = 2)
```

The following matrix shows food consumption after
removing the error.

Unlike weight, food consumption remained fairly 
steady over the course of the experiment. 

```{r remove_food_error}
outlier1 <- which(food.mat > 40)
food.mat[outlier1] <- NA
boxplot(food.mat, las = 2)
```

The food consumption matrix is also predominantly
in one dimention.

```{r food_decomp, warning = FALSE}
food.decomp <- plot.decomp(food.mat, plot.results = FALSE)
barplot_with_num(round(food.decomp$var.exp, 2), text.srt = 45)
```
This dimension is even more highly correlated with 
the average food intake, so we can safely reduce
food intake to its averge.

```{r food_pc}
plot.with.model(food.decomp$u[,1], food.mat[food.decomp$rows.used,"food_ave"],
  xlab = "PC1 of food intake", ylab = "Food Intake Average")
```

There is a weak relationship between food consumption and 
final weight.

```{r weight_v_food}
plot.with.model(food.mat[,"food_ave"], ordered.bw[,"weight_final"],
  xlab = "Avg. Food Consumption", ylab = "Weight (g)")
```

Some of the animals are crumblers, which means they store
food in their bedding. This will falsely increase the amount
of food they are consuming. However, I can't find any data 
to indicate which mice are crumblers and which are not.

The phenotypes fall into multiple categories, and some 
are normalized in different ways; i.e. adjusted for baseline
measurements or not, rank Z normalized or not. We don't want
to use all traits indiscriminately. Here we select traits 
for merging into a combined matrix that will capture multiple
dimensions of the metabolic state of the animals.

```{r add_food_consumption}
final.trait.mat <- cbind(final.trait.mat, food.mat[,"food_ave", drop=FALSE])
```

## Glucose

For glucose measurements we will look at glu_final, the 
final glucose measurement for each animal.

This measurement of glucose has the most variation.
It also captures the difference between animals whose glucose measurements
went up over the course of the experiment and those whose glucose
measurements went down over the course of the experiment. There is a 
slight negative correlation between initial glucose measurement and 
final glucose measurement. That is, animals that started with higher
glucose measurements tended to end with lower glucose measurements.

```{r glucose, fig.width = 8, fig.height = 8}

glu.mat <- adj.pheno[,c("glu_6", "glu_10", "glu_14", "glu_final")]
glu.mat[352,] <- NA
glu.diff <- glu.mat[,"glu_final"] - glu.mat[,"glu_6"] #did glucose go up or down over the course of the experiment?
glu.dir <- rep(NA, nrow(glu.mat))
glu.up <- which(glu.diff > 0) 
glu.down <- which(glu.diff < 0)
glu.dir[glu.up] <- "Up"
glu.dir[glu.down] <- "Down"

par(mfrow = c(2,2))
boxplot(glu.mat[,"glu_6"]~as.factor(glu.dir), xlab = "Glucose Change", 
  ylab = "Initial Glucose")
plot.with.model(glu.mat[,"glu_6"], glu.diff, report = "cor.test",
  xlab = "Initial Glucose", ylab = "Glucose Change")
plot.with.model(glu.mat[,"glu_final"], glu.diff, report = "cor.test",
  xlab = "Final Glucose", ylab = "Glucose Change")

```

```{r ins_hist, eval = FALSE}
sex.order <- order(as.numeric(as.factor(covar[,1])))
glu <- num.pheno[sex.order, "glu_final"]
weight <- num.pheno[sex.order, "weight_final"]
sex <- as.numeric(as.factor(covar[sex.order,1]))
not.outlier <- which(glu < 800)

pdf("~/Desktop/glucose_weight.pdf", height = 4, width = 4)
hist_with_points(glu[not.outlier], 
  col = sex[not.outlier]+2, xlab = "Fasting Glucose (mg/dL)",
  main = "", xaxis.at = seq(0, 400, 100), ylab = "Count")
  legend("topright", legend = c("Female", "Male"), pch = 16, col = unique(sex+2))
glucose.ranges <- c(99, 125) #from https://www.healthline.com/health/diabetes/fasting-glucose-normal-range#nondiabetic-levels
#abline(v = glucose.ranges)

plot.with.model(weight[not.outlier], glu[not.outlier], xlab = "Weight (g)",
  ylab = "Fasting Glucose (mg/dL)")

hist_with_points(num.pheno[sex.order,"weight_final"], col = sex+2,
  xlab = "Weight (g)", ylab = "Count", main = "")
  legend("topright", legend = c("Female", "Male"), pch = 16, col = unique(sex+2))
dev.off()
```


```{r add_final_glucose}
final.trait.mat <- cbind(final.trait.mat, adj.pheno[,"glu_final",drop=FALSE])
```

We also use the glucose total area under the curve. 
It is more highly correlated with final body weight
than is glu_iAUC.

```{r glucose_curves, fig.width = 8, fig.height = 8}
par(mfrow = c(2,2))
plot.with.model(adj.pheno[,"Glu_tAUC"], adj.pheno[,"Glu_iAUC"],
  xlab = "Glu_tAUC", ylab = "Glu_iAUC")
plot.with.model(adj.pheno[,"Glu_iAUC"], adj.pheno[,"weight_final"],
  xlab = "Glu_iAUC", ylab = "final weight")
plot.with.model(adj.pheno[,"Glu_tAUC"], adj.pheno[,"weight_final"],
  xlab = "Glu_tAUC", ylab = "final weight")
```

```{r add_glu_tAUC}
final.trait.mat <- cbind(final.trait.mat, adj.pheno[,"Glu_tAUC",drop=FALSE])
```

## Insulin 

Final insulin levels are also a good single measurement. 
This measurement captures the most variance, and is highly
correlated with the change in insulin over the course of the
experiment.

There is no correlation between initial insulin measurements
and the change in insulin over the course of the experiment.

```{r ins_levels, fig.width = 8, fig.height = 4}

ins.mat <- adj.pheno[,c("ins_6", "ins_10", "ins_14", "ins_final")]
ins.diff <- adj.pheno[,"ins_final"] - adj.pheno[,"ins_6"]

par(mfrow = c(1,2))
plot.with.model(ins.mat[,"ins_6"], ins.diff, xlab = "Initial Insulin", 
  ylab = "Insulin Change", report = "cor.test")
plot.with.model(ins.mat[,"ins_final"], ins.diff, xlab = "Final Insulin",
  ylab = "Insulin Change", report = "cor.test")
```


```{r add_ins_final}
final.trait.mat <- cbind(final.trait.mat, adj.pheno[,"ins_final",drop=FALSE])
```

For the insulin curves, we will select Ins_tAUC. 
This measurement is positively correlated with 
final body weight. The adjusted insulin AUC,
however, has a strange relationship with other traits
like final weight.

```{r ins_curves, fig.height = 4, fig.width = 8}

par(mfrow = c(1,2))
plot.with.model(adj.pheno[,"Ins_tAUC"], adj.pheno[,"weight_final"],
  xlab = "total insulin AUC", ylab = "final body weight", report = "cor.test")
plot.with.model(adj.pheno[,"Ins_iAUC"], adj.pheno[,"weight_final"],
  xlab = "adjusted insulin AUC", ylab = "final body weight", report = "cor.test")
```

```{r add_ins_tAUC}
final.trait.mat <- cbind(final.trait.mat, adj.pheno[,"Ins_tAUC", drop=FALSE])
```

## Triglycerides

There is an outlier in the first TG measurement at 6 weeks.
I'm going to remove it on suspicion that it is an error.

```{r tg}
tg.mat <- adj.pheno[,c("TG_6", "TG_10", "TG_14", "TG_final")]
boxplot(tg.mat)
outliers <- which(tg.mat > 600)
tg.mat[outliers] <- NA
```

The TG measurements have an interesting pattern: 
increasing early on from baseline, and then 
decreasing again.

```{r tg_final}
boxplot(tg.mat)
```

There also isn't one dimenson of variation.
Should we use all TG measurements? The first PC?

```{r tg_decomp}
tg.decomp <- plot.decomp(tg.mat, plot.results = FALSE)
barplot_with_num(round(tg.decomp$var.exp, 2), ylab = "Variance Explained")
```

```{r add_tg}
final.trait.mat <- cbind(final.trait.mat, tg.mat)
```


## Pancreas traits and HOMA

We selected all three pancreas traits: WPIC, 
Ins_per_islet, and num_islets. And both HOMA
traits

```{r add_panc_traits}
final.trait.mat <- cbind(final.trait.mat, 
  adj.pheno[,c("WPIC", "Ins_per_islet", "num_islets", "HOMA_B", "HOMA_IR")])
```

## Cholesterol

The cholesterol measurements are highly correlated.


```{r chol}
chol.traits <- c("HDL_3_2a", "HDL_2b", "Mid_Zone", "LDL_IVc", 
  "LDL_IVb", "LDL_IVa", "LDL_IIIb", "LDL_IIIa", "LDL_IIb", 
  "LDL_IIa", "LDL_I", "IDL_2", "IDL_1", "VLDL_sm", "VLDL_int", 
  "VLDL_lg", "LDL_pk_diam")

chol.mat <- adj.pheno[,chol.traits]
keep.rows <- which(apply(chol.mat, 1, function(x) !all(is.na(x))))
chol.mat <- chol.mat[keep.rows,]
pheatmap(cor(chol.mat))
```

There are two major axes of variation in this matrix,
but eight total axes of variation across the individuals. 

```{r chol_decomp, fig.width = 8, fig.height = 4, warning = FALSE}
chol.decomp <- plot.decomp(chol.mat, plot.results = FALSE, pc = 8)
barplot_with_num(round(chol.decomp$var.exp, 2), main = "PC Variance Explained",
  ylab = "Variance Explained", text.srt = 45)
```

The PCs of the cholesterol matrix are correlated with the
original measures in the following way. 
The first PC captures variance mostly in the LDL and IDL measurements. 
The second PC captures variance in the VLDL measurements. 
The third PC is negatively correlated with LDL_pd_diam.
The fourth PC is negatively correlated with HDL.

It looks as if the first four PCs capture the variance
that is translatable to the traits. We will use the first
four PCs of this matrix. 

```{r chol_pc_cor}

chol.pc <- chol.decomp$u
pc.cor.mat <- matrix(NA, nrow = ncol(chol.pc), ncol = ncol(chol.mat))
colnames(pc.cor.mat) <- colnames(chol.mat)
rownames(pc.cor.mat) <- paste0("PC", 1:nrow(pc.cor.mat))
for(i in 1:ncol(chol.pc)){
  pc.cor <- apply(chol.mat, 2, function(x) cor(x, chol.pc[,i], 
    use = "pairwise.complete.obs"))
  pc.cor.mat[i,] <- pc.cor
}
pheatmap(t(pc.cor.mat), cluster_cols = FALSE)

chol.pc.mat <- chol.decomp$u[,1:4]
rownames(chol.pc.mat) <- rownames(chol.mat)
colnames(chol.pc.mat) <- paste0("Chol_PC", 1:ncol(chol.pc.mat))

excluded.names <- setdiff(rownames(final.trait.mat), rownames(chol.pc.mat))
chol.pc.padding <- matrix(NA, nrow = length(excluded.names), ncol = ncol(chol.pc.mat))
rownames(chol.pc.padding) <- excluded.names
colnames(chol.pc.padding) <- colnames(chol.pc.mat)
full.pc.mat <- rbind(chol.pc.padding, chol.pc.mat)
row.order <- match(rownames(final.trait.mat), rownames(full.pc.mat))
final.pc.mat <- full.pc.mat[row.order,]
#identical(rownames(final.pc.mat), rownames(final.trait.mat))
```

```{r add_chol_pc}
final.trait.mat <- cbind(final.trait.mat, final.pc.mat)
```

## The rest

We added:
* the rank Z normalized glycogen measurement.
* the rank Z normalized proinsulin measurement.
* Gcg content
* Gcg secreted
* tibia bone mineral density (BMD)
* leptin and adiponectin levels
* organ weights (heart, liver, fat pad)
* adiposity

```{r add_the_rest}
final.trait.mat <- cbind(final.trait.mat, 
  adj.pheno[,c("glycogen_rz", "proinsulin_rz", "Gcg_content", "Gcg_secreted",
  "BMD", "leptin", "adiponectin", "heart_weight", "liver_weight",
  "fat_pad_weight", "adiposity")])
```

The following barplot shows the number of individuals
with measurements for each trait. Some of them don't have
many individuals.

```{r ind_count, fig.width = 8, fig.height = 5}
ind.count <- apply(final.trait.mat, 2, function(x) length(which(!is.na(x))))
par(mar = c(8, 4, 4, 4))
barplot_with_num(ind.count, las = 2, text.srt = 45, text.gap = 0.2, 
  text.shift = 0.1)

total.ind <- length(which(!is.na(rowSums(final.trait.mat))))
```

If we use the full matrix, there are only `r total.ind` individuals
with all traits measured. Unfortunately we'll have to 
pare down even from here which traits to use. 

The following heatmap shows where blocks of missing 
individuals are. We will remove the organ weights, 
leptin, adiponectin, and glucagon, and adiposity. 

```{r remove_which}
remove.traits <- c("Gcg_content", "Gcg_secreted", "adiponectin",
  "leptin", "heart_weight", "liver_weight", "fat_pad_weight", "adiposity")

is.included <- rep("yes", ncol(final.trait.mat))
not.included <- which(colnames(final.trait.mat) %in% remove.traits)
is.included[not.included] <- "no"
included <- data.frame("included" = is.included)
rownames(included) <- colnames(final.trait.mat)
```

```{r missing_ind, fig.width = 8, fig.height = 6}
pheatmap(final.trait.mat, cluster_rows = FALSE, cluster_cols = FALSE,
  show_rownames = FALSE, annotation_col = included)
```

Before we remove the traits with lots of missing values, we
looked at the correlation matrix for these traits to compare
with the PC matrix below. I used Spearman rank correlation
because some of these traits have really weird distributions.

```{r trait_cor, fig.width = 9, fig.height = 7}
full.cor <- cor(final.trait.mat, use = "pairwise.complete.obs", method = "spearman")
pheatmap(full.cor, annotation_row = included)
```

The following heatmap shows the correlation matrix
only for the traits that are included in the final 
trait set.

```{r remove_traits, warning = FALSE, fig.width = 7, fig.height = 7}
keep.traits <- setdiff(colnames(final.trait.mat), remove.traits)
pared.traits <- final.trait.mat[,keep.traits]

norm.traits <- apply(pared.traits, 2, rankZ)
norm.cor <- cor(norm.traits, use = "pairwise.complete.obs")
pheatmap(norm.cor, cellwidth = 12, cellheight = 12)

imageWithTextColorbar(norm.cor, use.pheatmap.colors = TRUE,
  orientation = "h", cex = 1, bar.lwd = 2, round.nearest = 0.1)


norm.decomp <- plot.decomp(norm.traits, pc = ncol(norm.traits), plot.results = FALSE)
#barplot_with_num(round(norm.decomp$var.exp, 2))

```

This trait set includes `r nrow(norm.decomp$u)` animals.
Some of the excluded traits are correlated with the first PC of this
matrix. The organ weights and leptin levels are correlated with 
insulin_tAUC and HOMA_IR. However, adiposity, adiponectin, and 
glucagon are not correlated with anything else in this matrix.

I'm not sure how adiposity was measured. It doesn't have an 
explanation in the phenotype annotation matrix. Although fat
pad weight and body weight are highly correlated, neither
is correlated with adiposity, which makes me suspicious of that
measurement.

Why are those cholesterol PCs dominating some of the PCs?
Is that a problem? Is it interesting?


```{r pc_v_traits, fig.width = 7, fig.height = 7}
trait.cor.mat <- matrix(NA, nrow = ncol(norm.decomp$u), 
  ncol = ncol(final.trait.mat))
colnames(trait.cor.mat) <- colnames(final.trait.mat)
rownames(trait.cor.mat) <- paste0("PC", 1:nrow(trait.cor.mat))

for(i in 1:ncol(norm.decomp$u)){ #for each PC
  trait.cor <- apply(final.trait.mat[norm.decomp$rows.used,], 2,
    function(x) cor(x, norm.decomp$u[,i], use = "pairwise.complete.obs", method = "spearman"))
  trait.cor.mat[i,] <- trait.cor
}

pheatmap(t(trait.cor.mat), cluster_cols = FALSE, annotation_row = included)
```

The variance explained by each of these PCs is the following.

```{r var_exp, fig.width = 7, fig.height = 5}
var.exp <- round(norm.decomp$var.exp, 2)
names(var.exp) <- paste("PC", 1:length(var.exp))
barplot_with_num(var.exp, las = 2, text.srt = 45)
```

```{r test, eval = FALSE}
pc.num = 1
trait.name = "leptin"
plot.with.model(rankZ(norm.decomp$u[,pc.num]), 
  rankZ(final.trait.mat[norm.decomp$rows.used,trait.name]), report = "cor.test",
  xlab = paste("PC", pc.num), ylab = trait.name)

plot.with.model(rankZ(final.trait.mat[,"fat_pad_weight"]), 
  rankZ(final.trait.mat[,"weight_final"]), report = "cor.test")

```

The final phenotype matrix is stored in the Data folder in the file
Clinical_Phenotypes_Filtered.RDS

```{r write_data}
saveRDS(pared.traits, here("Data", "DO", "Clinical_Phenotypes_Adjusted.RDS"))
```

