---
title: "Mouse Validation"
author: Anna L Tyler
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
  html_document:
    code_folding: hide
    collapsed: no
    toc: yes
    toc_float: yes
---

## Introduction

The purpose of this workflow is to validate the transcript
loadings we identified in HDMA in another population. Rather
than do this in the CC-RIX, which would be the first publication
for those data, we can use a published data set, like the Svenson
DO. The Svenson DO only measured liver expression, so if this
doesn't work, it's not a big deal. 

```{r load_code}
rm(list = ls())

#is.interactive = TRUE
library("here")

exp.name = "tissue_together-_-complete_mediation-germline_kinship"
data.results.dir <- here("Results", "High_Dim_Med", exp.name)

all.fun <- list.files(here("Code"), full.names = TRUE, pattern = ".R")
for(i in 1:length(all.fun)){source(all.fun[i])}
```

```{r load_libraries, message = FALSE, warning = FALSE, error = FALSE}
all.packages <- c("pheatmap", "gprofiler2", "RColorBrewer", "GEOquery", "limma")
load_libraries(all.packages, personal.library = TRUE)
```


```{r read_mouse_data}
transcript_loadings <- readRDS(file.path(data.results.dir, "Loadings_Transcripts_0.RDS"))
tissue.names <- names(transcript_loadings)

svenson_do <- load(here("Data", "DO", "Svenson_DO850_for_eQTL_viewer_v9.RData"))
norm_expr <- dataset.mrna$data$norm

covar.mat <- dataset.phenotype$covar.matrix

#subset by diet
hfd.idx <- which(covar.mat[,"diethf"] == 1)
chow.idx <- which(covar.mat[,"diethf"] == 0)

common.hfd <- intersect(names(hfd.idx), rownames(norm_expr))
common.chow <- intersect(names(chow.idx), rownames(norm_expr))

hfd.expr <- norm_expr[common.hfd,]
chow.expr <- norm_expr[common.chow,]

#adjust for sex and generation
adj_hfd <- adjust(hfd.expr, covar.mat[,c(1,3:8)])
adj_chow <- adjust(chow.expr, covar.mat[,c(1,3:8)])

scaled_hfd <- apply(adj_hfd, 2, scale)
dimnames(scaled_hfd) <- dimnames(adj_hfd)

scaled_chow <- apply(adj_chow, 2, scale)
dimnames(scaled_chow) <- dimnames(adj_chow)
```

```{r pheno}
pheno <- dataset.phenotype$data
hfd.adj.pheno <- adjust(pheno[common.hfd,], covar.mat[,c(1,3:8)])
chow.adj.pheno <- adjust(pheno[common.chow,], covar.mat[,c(1,3:8)])
```

```{r load_expression, fig.width = 8, fig.height = 8}
#pdf("~/Desktop/DO850.pdf")
par(mfrow = c(2,2))
for(tx in 1:length(tissue.names)){
    tx.loadings <- transcript_loadings[[tx]]
    common.tx <- intersect(colnames(scaled_hfd), rownames(tx.loadings))
    loaded_chow <- apply(scaled_chow[,common.tx], 1, function(x) x*tx.loadings[common.tx,])
    loaded_hfd <- apply(scaled_hfd[,common.tx], 1, function(x) x*tx.loadings[common.tx,])

    pred.chow <- colMeans(loaded_chow)
    pred.hfd <- colMeans(loaded_hfd)

    common.hfd <- intersect(names(pred.hfd), rownames(hfd.adj.pheno))
    plot.with.model(pred.hfd[common.hfd], hfd.adj.pheno[common.hfd,"bw_pc1"], 
        xlab = "Prediction", ylab = "BW PC1", main = paste("HFD", tissue.names[tx]))

    common.chow <- intersect(names(pred.chow), rownames(chow.adj.pheno))
    plot.with.model(pred.chow[common.chow], chow.adj.pheno[common.chow,"bw_pc1"], 
        xlab = "Prediction", ylab = "BW PC1", main = paste("Chow", tissue.names[tx]))

}
#dev.off()

```

```{r all_cor}
chow.cor <- hfd.cor <- matrix(NA, nrow = length(tissue.names), ncol = ncol(pheno))
rownames(chow.cor) <- rownames(hfd.cor) <- tissue.names
colnames(chow.cor) <- colnames(hfd.cor) <- colnames(pheno)

for(tx in 1:length(tissue.names)){
    tx.loadings <- transcript_loadings[[tx]]
    common.tx <- intersect(colnames(adj_hfd), rownames(tx.loadings))
    loaded_chow <- apply(scaled_chow[,common.tx], 1, function(x) x*tx.loadings[common.tx,])
    loaded_hfd <- apply(scaled_hfd[,common.tx], 1, function(x) x*tx.loadings[common.tx,])

    pred.chow <- colMeans(loaded_chow)
    pred.hfd <- colMeans(loaded_hfd)

    common.chow <- intersect(names(pred.chow), rownames(chow.adj.pheno))
    all.chow.cor <- apply(chow.adj.pheno[common.chow,], 2, 
        function(x) cor(x, pred.chow[common.chow], use = "pairwise.complete.obs", 
        method = "spearman"))
    chow.cor[tx,] <- all.chow.cor

    common.hfd <- intersect(names(pred.hfd), rownames(hfd.adj.pheno))
    all.hfd.cor <- apply(hfd.adj.pheno[common.hfd,], 2, 
        function(x) cor(x, pred.hfd[common.hfd], use = "pairwise.complete.obs", 
        method = "spearman"))
    hfd.cor[tx,] <- all.hfd.cor
}
```

The following heat map shows that the liver loadings give 
us the highest correlations with body weight and insulin
measurements. This heat map shows results for the chow 
animals only.

```{r chow_trait_cor, fig.width = 4, fig.height = 11}
pheatmap(t(chow.cor), cluster_cols = FALSE)
```

The following heat map is the same as above, but is
for the HFD animals only.


```{r hfd_trait_cor, fig.width = 4, fig.height = 11}
pheatmap(t(hfd.cor), cluster_cols = FALSE)
```