I'm taking out all the digging into distal eQTL.
I've learned from Gary that this is a minefield of pseudogenes and 
chromosome-level variation.

## Aligning Distal eQTL Hotspots {.tabset .tabset-fade .tabset-pills}

The following plots show alignments of distal eQTL hotspots
across the different tissues. There are quite a few multi-tissue
hotspots. We investigate these in more detail below.

```{r align_hotspots, fig.height = 5, fig.width = 10, results = "asis"}
distal.targets.by.chr <- vector(mode = "list", length = 20)
for(tx in 1:length(distal.target.list)){
  target.counts <- sapply(distal.target.list[[tx]], function(x) sapply(x, length))
  for(ch in 1:length(distal.targets.by.chr)){
    distal.targets.by.chr[[ch]] <- rbind(distal.targets.by.chr[[ch]], target.counts[[ch]])
  }
}

#pdf("~/Desktop/hotspots_no_pseudo.pdf", width = 10, height = 5)
for(ch in 1:20){
  cat("### Chr", ch, "\n")
  layout(matrix(c(1,2), nrow = 1), widths = c(1, 0.2))
  combined.targets <- distal.targets.by.chr[[ch]]
  rownames(combined.targets) <- tissue.names
  coord <- as.numeric(colnames(combined.targets))
  par(mar = c(4,4,4,0))
  plot.new()
  plot.window(xlim = c(0, ceiling(max(coord)/50)*50), ylim = c(0, max(combined.targets)))
  for(tx in 1:nrow(combined.targets)){
    points(coord, combined.targets[tx,], type = "l", col = tissue.cols[tx], lwd = 3)
  }
  axis(1)
  axis(2)
  mtext(paste("Chromosome", ch, "Position (Mb)"), side = 1, line = 2.5)
  mtext("Target Count", side = 2, line = 2.5)
  mtext(paste("Chromosome", ch), side = 3)
  par(mar = c(4,0,4,0))
  plot.new()
  plot.window(xlim = c(0,1), ylim = c(0,1))
  legend(x = 0, y = 1, legend = tissue.names, col = tissue.cols, lty = 1, 
    lwd = 3, adj = 0)
  cat("\n\n")
}
#dev.off()
```

## Distal hotspot expression {.tabset .tabset-fade .tabset-pills}

In the following plots we show multiple aspects of these
distal eQTL hotspots. We show the following items for each
hotspot:

1. The first two PCs of the expression matrix of the target
  genes in terms of transcripts. Transcripts are colored
  by the tissue they are expressed in. For some hotspots
  the transcripts cluster by tissue, and in others they don't.

2. The first two PCs of the expression matrix of the target
  genes in terms of individuals. Clustering of individuals tends
  to correspond with high LOD scores for the first PC of the 
  expression matrix.

3. A LOD trace for the first PC of the target gene expression
  matrix.

4. The haplotype coefficients for the first PC of the target
  gene expression matrix.

5. A list of target genes for this eQTL split by tissue.

6. Functional enrichment plots for the target genes if they
  have any functional enrichment.


decomposition of the 
expression of targets of these hotspots both in terms of
transcripts and in terms of individuals. Tight clustering
of individuals corresponds with strong eQTL LOD scores for
the first principal component of the gene expression of the 
whole group of targets. 


What does expression look like for groups of transcripts
with shared distal eQTL across tissues? 

Unfortunately, I
have found that the targets of these hotspots tend to be predicted
and pseudogenes. If the target genes contain many protein-coding 
genes, there tends not to be an eQTL for the expression of the 
target genes as a whole. So multi-tissue eQTL hotspots are
unlikely to be related to real biology, they are just drift.


```{r pos_candidate_expression}
get_region_info <- function(chr, min.mbp, max.mbp){
  chr.tables <- lapply(tissue.data, function(x) x$annot.mrna[which(x$annot.mrna$chr == chr),])
  in_region <- lapply(chr.tables, function(x) intersect(which(x$start >= min.mbp), which(x$end <= max.mbp)))
  region.tables <- lapply(1:length(chr.tables), function(x) chr.tables[[x]][in_region[[x]],])
  return(region.tables)
}

get_region_expr <- function(chr, min.mbp, max.mbp){
  region.tables <- get_region_info(chr, min.mbp, max.mbp)
  common.id <- lapply(1:length(region.tables), function(x) intersect(region.tables[[x]]$gene.id, colnames(ind.matched.expr[[x]])))
  region.expr <- lapply(1:length(region.tables), function(x) ind.matched.expr[[x]][,common.id[[x]]])
  expr.names <- unlist(lapply(1:length(region.expr), function(x) paste(colnames(region.expr[[x]]), tissue.names[x], sep = "_")))
  merged.region.expr <- Reduce("cbind", region.expr)
  colnames(merged.region.expr) <- expr.names
  return(merged.region.expr)
}

```

```{r fig.width = 12, fig.height = 9, results = "asis", error = FALSE, warning = FALSE, message = FALSE}

#pdf("~/Desktop/hotspot_decomp.pdf", width = 12, height = 9)
min.thresh <- 10
n.hot <- 0
for(ch in 1:20){
  cat("### Ch", ch, "{.tabset .tabset-fade .tabset-pills}\n")
  combined.targets <- distal.targets.by.chr[[ch]]
  rownames(combined.targets) <- tissue.names
  shared.hotspot <- which(apply(combined.targets, 2, function(x) length(which(x >= min.thresh))) > 1)
  #n.hot <- n.hot + length(shared.hotspot)

  if(length(shared.hotspot) == 0){next()}
  for(h in 1:length(shared.hotspot)){
    cat("####", paste("Chromomsome", ch, names(shared.hotspot)[h], "Mb"), "\n")
    tx.transcripts <- lapply(distal.target.list, function(x) x[[ch]][match(names(shared.hotspot)[h], names(x[[ch]]))])
    target.expr <- lapply(1:length(tx.transcripts), function(x) ind.matched.expr[[x]][,unlist(tx.transcripts[[x]]),drop=FALSE])
    target.info <- lapply(1:length(target.expr), function(x) tissue.data[[x]]$annot.mrna[match(colnames(target.expr[[x]]), tissue.data[[x]]$annot.mrna$gene.id),])
    
    expr.names <- unlist(lapply(1:length(target.expr), function(x) if(length(target.expr[[x]]) > 0){paste(colnames(target.expr[[x]]), tissue.names[x], sep = "_")}))
    merged.expr <- Reduce("cbind", target.expr)
    colnames(merged.expr) <- expr.names
    combined.hap.effects <- lapply(1:8, function(x) sapply(expr.scan, function(y) y[,x]))
    expr.col <- unlist(lapply(1:length(target.expr), function(x) rep(tissue.cols[x], ncol(target.expr[[x]]))))
    #pheatmap(cor(merged.expr))
    layout(matrix(c(1,2,3,6,4,4,4,7,5,5,5,8), nrow = 3, byrow = TRUE), widths = c(1,0.2,1,2))
    par(mar = c(4,4,4,0))
    plot.decomp(cor(merged.expr), main = "Transcripts", cols = expr.col)
    par(mar = c(4,0,4,0))
    plot.new()
    plot.window(xlim = c(0,1), ylim = c(0,1))
    legend(x = 0, y = 1, legend = tissue.names, col = tissue.cols, pch = 16)
    par(mar = c(4,4,4,4))
    expr.decomp <- plot.decomp(merged.expr, main = "Individuals")
    expr.pc <- expr.decomp$u
    rownames(expr.pc) <- rownames(merged.expr)[expr.decomp$rows.used]
    decomp.scan <- scan1coef(genoprobs[,ch], expr.pc[,1])
    decomp.lod <- scan1(genoprobs[,ch], expr.pc[,1])
    peak.pos <- find_peaks(decomp.lod, map = map, prob = 0.9)

    ##====================================================
    ## plot mapping results
    ##====================================================
    par(mar = c(4,4,4,4))
    plot(decomp.lod, map = map)
    plot.lim <- par("usr")
    segments(x0 = as.numeric(names(shared.hotspot)[h]), 
      y0 = plot.lim[3], y1 = plot.lim[4], lwd = 2)
    plot_coefCC(decomp.scan, map = map, main = "PC1 coefficients")
    plot.lim <- par("usr")
    segments(x0 = as.numeric(names(shared.hotspot)[h]), 
      y0 = plot.lim[3], y1 = plot.lim[4], lwd = 2)

    ##====================================================
    ## enrichment
    ##====================================================
    mtext(paste("Chromomsome", ch, names(shared.hotspot)[h], "Mb"), outer = TRUE, line = -1.5)
    enrich <- gost(unique(unlist(tx.transcripts)), organism = "mmusculus")
    target.list <- lapply(target.info, function(x) words_to_paragraph(x$symbol))
    gene.text <- unlist(lapply(1:length(target.list), 
      function(x) paste(paste(tissue.names[x], target.list[[x]], sep = ": "), "\n")))
    par(mar = c(0,0,0,0))
    plot.text("Target Genes", y = 1)
    plot.text(paste(gene.text, collapse = "\n"), add = TRUE)
    plot.enrichment.wordcloud(enrich,  num.terms = 20, order.by = "p_value", 
    just.wordcloud = TRUE)  


    ##====================================================
    ## mediation
    ##====================================================

    mediation.file <- here("Results",paste0("Mediation_Results_Chr", ch, "_", names(shared.hotspot)[h], "Mb.txt"))

    if(!file.exists(mediation.file)){
      candidate.info <- get_region_info(ch, peak.pos[1,"ci_lo"]-1, peak.pos[1,"ci_hi"]+1)
      candidate.expr <- get_region_expr(ch, peak.pos[1,"ci_lo"]-1, peak.pos[1,"ci_hi"]+1)
      peak.marker <- find_marker(map, ch, peak.pos[,"pos"])
      peak.geno <- genoprobs[[ch]][,,peak.marker]
      med.test <- bmediatR(y = expr.pc[,1], M = candidate.expr, X = peak.geno,
        ln_prior_c = "reactive", options_X = list(sum_to_zero = TRUE, 
        center = FALSE, scale = FALSE))
      #plot_posterior_bar(med.test, mediator_id = colnames(candidate.expr)[1])
      post.mat <- get_posterior(med.test, med_var = "transcript.id")
      write.table(post.mat, mediation.file, sep = "\t", quote = FALSE, row.names = FALSE)
    }else{
      post.mat <- read.table(mediation.file, sep = "\t", header = TRUE)
    }

    interesting.mat <- as.matrix(post.mat[,c(5, 9, 12, 13)])
    rownames(interesting.mat) <- post.mat[,1]

    min.region <- min(sapply(candidate.info, function(x) min(x$start)))
    max.region <- max(sapply(candidate.info, function(x) max(x$start)))

    split.tx <- strsplit(colnames(candidate.expr), "_")
    just.tx <- sapply(split.tx, function(x) x[1])  
    just.tissue <- sapply(split.tx, function(x) x[2])
    info.idx <- sapply(1:length(just.tx), function(x) which(tissue.names == just.tissue[x]))
    tx.pos <- unlist(sapply(1:length(just.tx), function(x) candidate.info[[info.idx[x]]][which(candidate.info[[info.idx[x]]]$gene.id == just.tx[x]),"start"]))

    par(mar = c(4,4,4,4))
    plot.new()
    plot.window(xlim = c(min.region-1, max.region), ylim = c(0.5, 5))
    for(i in 1:ncol(interesting.mat)){
      segments(x0 = tx.pos, y0 = rep(i, length(tx.pos)), 
        y1 = rep(i, length(tx.pos))+interesting.mat[,i]*0.9, lwd = 2, 
        col = tissue.cols[sapply(just.tissue, function(x) which(tissue.names == x))])
      text(x = min.region-0.15, y = i+0.5, labels = colnames(interesting.mat)[i], adj = 1)
      draw.rectangle(min.region-0.1, max.region, i-0.05, i+0.9)
    }  
    ax.ticks <- pretty(floor(min.region):ceiling(max.region), n = 5)
    show.which <- intersect(which(ax.ticks >= min.region-0.1), which(ax.ticks <= max.region))
    segments(x0 = ax.ticks[show.which], y0 = 0.95, y1 = 0.85)
    text(x = ax.ticks[show.which], y = 0.75, labels = ax.ticks[show.which])
    text(x = mean(c(min.region, max.region)), y = 0.45, paste("Chromosome", ch, "Position (Mb)"))
    #legend("bottomleft", lty = 1, lwd = 3, col = tissue.cols, legend = tissue.names)
    text(x = mean(c(min.region, max.region)), y = 5.1, labels = "Mediation Results")
    ##cat(unique(colnames(merged.expr)), sep = "\n")

    ##====================================================
    cat("\n\n")
  }
  cat("\n\n")
}
#dev.off()
```
## Distal eQTL Hotspot Enrichments {.tabset .tabset-fade .tabset-pills}

We looked for functional enrichments in distal targets of 
eQTL hotspots defined above.

```{r hotspot_enrich, fig.height = 12, fig.width = 9, message = FALSE, error = FALSE, warning = FALSE, results = "asis"}

for(tx in 1:length(tissue.names)){
  cat("###", tissue.names[tx], "{.tabset .tabset-fade .tabset-pills}\n")
  tx.distal <- distal.target.list[[tx]]
  tx.distal.counts <- lapply(tx.distal, function(x) sapply(x, length))
  hotspot.idx <- lapply(tx.distal.counts, function(x) which(x >= hotspot.min))  
  hotspot.names <- lapply(1:length(hotspot.idx), function(x) if(length(hotspot.idx[[x]]) > 0){tx.distal[[x]][hotspot.idx[[x]]]})
  
  #u_hotspot.transcripts <- unique(unlist(hotspot.names))
  #eqtl.chr <- sapply(u_hotspot.transcripts, function(x) grep(x, hotspot.names)[1])
  #hotspot.expr  <- adj.expr[[tx]][,u_hotspot.transcripts]
  #expr.decomp <- plot.decomp(cor(hotspot.expr), plot.results = FALSE)
  #plot(expr.decomp$u, type = "n", xlab = "PC1", ylab = "PC2", 
  #  main = paste(tissue.names[tx], "distal eQTL target expression decomposition"))
  #text(expr.decomp$u[,1], expr.decomp$u[,2], col = eqtl.chr, labels = eqtl.chr)

  hotspot.enrich <- lapply(hotspot.names, function(x) lapply(x, function(y) gost(y, organism = "mmusculus")))
  for(i in 1:length(hotspot.enrich)){
    if(length(hotspot.enrich[[i]])  > 0){
      names(hotspot.enrich[[i]]) <- paste0("Chr", i, "_", names(hotspot.enrich[[i]]), "Mb")
    }
  }
  enrich.list <- unlist(hotspot.enrich, recursive = FALSE)  
  if(length(unlist(enrich.list)) == 0){next()}
  #pdf(paste0("~/Desktop/enrich_", tissue.names[tx], ".pdf"), height = 14, width = 9)
  plot.enrichment.group(enrich.list, transformation = "sqrt", 
    plot.label = paste("Distal eQTL targets in", tissue.names[tx]),
    cluster_cols = FALSE)
  #dev.off()
  cat("\n\n")
}

```
