---
title: "GSEA"
author: Anna L Tyler
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
  html_document:
    code_folding: hide
    collapsed: no
    toc: yes
    toc_float: yes
---

## Introduction

The purpose of this workflow is to use GSEA to look for 
enrichments in the transcript loading lists as an alternative
to the gProfiler method of enrichment. With GSEA there is no 
need to use a cutoff and look only at the top-ranked genes.

Here we look at both GO and KEGG gene sets.

```{r load_code}
rm(list = ls())

is.interactive = FALSE
#is.interactive = TRUE

library("here")

all.fun <- list.files(here("Code"), full.names = TRUE, pattern = ".R")
for(i in 1:length(all.fun)){source(all.fun[i])}

kegg.result.dir <- here("Results", "KEGG")
if(!file.exists(kegg.result.dir)){dir.create(kegg.result.dir, recursive = TRUE)}

go.result.dir <- here("Results", "GO")
if(!file.exists(kegg.result.dir)){dir.create(kegg.result.dir, recursive = TRUE)}
```

```{r load_libraries, message = FALSE, warning = FALSE, error = FALSE}
all.packages <- c("clusterProfiler", "fgsea", "pheatmap", "gprofiler2", "pathview", 
    "vioplot", "pRoloc", "annotate", "GO.db")
load_libraries(all.packages)
```

```{r load_hdma_results}
tissue.cols <- as.matrix(read.delim(here("Data", "general", "tissue_colors.txt"), header = FALSE, row.names = 1))[,1]

transcript_loadings <- readRDS(here("Results", "DO", "High_Dim_Med_no_MGE", 
    "tissue_together-_-complete_mediation-germline_kinship", "Loadings_Transcripts_0.RDS"))
tissue.names <- names(transcript_loadings)
```

## KEGG enrichments

We used the package clusterProfiler to download KEGG 
annotations for mouse genes.

```{r get_kegg}
#download and read in mouse KEGG annotations using download_KEGG()
#from the package clusterProfiler
kegg.file <- here("Data", "general", "KEGG.Mouse.RDS")
if(!file.exists(kegg.file)){
    all.kegg <- download_KEGG("mmu", "KEGG", "kegg")
    saveRDS(all.kegg, kegg.file)
}else{
    all.kegg <- readRDS(kegg.file)
}

u_path <- gsub(" - Mus musculus (house mouse)", "", all.kegg[[2]][,"to"], fixed = TRUE)

#load gene conversion table
gene.table <- read.table(here("Data", "general", "mouse_gene_info.txt"), sep = "\t", header = TRUE)
```

We biult a list of KEGG terms and their gene assignments.

```{r build_list}
# build GSEA list
path.locale <- lapply(u_path, function(x) grep(x, all.kegg[[2]][,2], fixed = TRUE))
path.id <- lapply(path.locale, function(x) all.kegg[[2]][x,][1])
path.idx <- lapply(path.id, function(x) which(all.kegg[[1]][,1] == unlist(x)[1]))
path.gene.id <- lapply(path.idx, function(x) all.kegg[[1]][x,2])
names(path.gene.id) <- u_path

path.gene.ensembl <- lapply(path.gene.id, 
    function(x) gene.table[match(x, gene.table[,"entrezgene_id"]), "ensembl_gene_id"])
```

We then used fGSEA to calculate normalized enrichment scores (NES) 
for each KEGG pathway using the transcript loadings from each of 
the four tissues.

```{r calc_enrich_kegg}
path_enrich <- matrix(NA, nrow = length(u_path), ncol = length(tissue.names))
rownames(path_enrich) <- u_path
colnames(path_enrich) <- tissue.names

all_min <- min(sapply(transcript_loadings, min))
all_max <- max(sapply(transcript_loadings, max))

path_enrich_file <- file.path(kegg.result.dir, "KEGG_pathway_enrichments.RDS")

if(!file.exists(path_enrich_file)){
    #get enrichment scores for all pathways and all tissues
    for(tx in 1:length(tissue.names)){
        sorted.tx.loadings <- transcript_loadings[[tx]][order(transcript_loadings[[tx]], decreasing = TRUE),1]
        gsea.enrich <- fgsea::fgsea(path.gene.ensembl, sorted.tx.loadings, scoreType = "std")
        norm.es <- as.numeric(as.matrix(gsea.enrich[,"NES"]))
        names(norm.es) <- gsea.enrich$pathway

        name.idx <- match(names(norm.es), rownames(path_enrich))
        path_enrich[name.idx,tx] <- norm.es
    }
    saveRDS(path_enrich, path_enrich_file)
}else{
    path_enrich <- readRDS(path_enrich_file)
}
```

### NES Distributions

The distributions of the NES are shown below.

```{r nes_dist}
vioplot(path_enrich, col = tissue.cols, ylab = "GSEA normalized enrichment score")
abline(h = 0)
```

### Top KEGG Enrichments {.tabset .tabset-fade .tabset-pills}

The top and bottom 10 enrichment scores are shown below.

```{r plot_nes, warning = FALSE, message = FALSE, error = FALSE, results = "asis", fig.width = 9, fig.height = 7}
#pdf(file.path(kegg.result.dir, "NES_by_tissue.pdf"), width = 9, height = 7)
top.scores <- vector(mode = "list", length = length(tissue.names))
names(top.scores) <- tissue.names

par(mar = c(4,4, 2, 4))
for(tx in 1:length(tissue.names)){
    cat("####", tissue.names[tx], "\n")
    sorted.enrich <- path_enrich[order(path_enrich[,tx], decreasing = FALSE),tx]
    extreme.enrich  <- c(head(sorted.enrich, 10), tail(sorted.enrich, 10))
    a <- barplot(extreme.enrich, 
        horiz = TRUE, las = 2, main = tissue.names[tx], 
        xlim = c(min(path_enrich), max(path_enrich)), name = NA)
    text(x = rep(-0.1, 10), y = tail(a, 10), labels = tail(names(extreme.enrich), 10), adj = 1)
    text(x = rep(0.1, 10), y = head(a, 10), labels = head(names(extreme.enrich), 10), adj = 0)
    mtext("Normalized Enrichment Score", side = 1, line = 2.5)
    top.scores[[tx]] <- names(extreme.enrich)
    cat("\n\n")
}
#dev.off()
```

### Loadings for KEGG pathways

Plot the loading distributions for the most extreme pathways. 
The p value above each distribution is from a t test testing 
whether the mean loading is non-zero.

```{r plot_all_dist, warning = FALSE, message = FALSE, error = FALSE, fig.width = 9, fig.height = 9}
all.extreme.kegg <- unique(unlist(top.scores))
top.kegg.mat <- path_enrich[match(all.extreme.kegg, rownames(path_enrich)),]

kegg.loading.file <- file.path(kegg.result.dir, "Kegg_Pathway_Loadings.pdf")

if(!file.exists(kegg.loading.file)){
    pdf(kegg.loading.file, width = 12, height = 12)
    par(mfrow = c(4,4))
    for(i in 1:length(all.extreme.kegg)){
        if(is.interactive){
            report.progress(i, length(u_path))
        }
        search.term <- all.extreme.kegg[i]

        path.locale <- grep(search.term, all.kegg[[2]][,2], fixed = TRUE)
        path.id <- all.kegg[[2]][path.locale,][1]
        path.name <- all.kegg[[2]][path.locale,][2]

        path.idx <- which(all.kegg[[1]][,1] == unlist(path.id))
        path.gene.id <- all.kegg[[1]][path.idx,2]
        path.gene.table <- gconvert(as.numeric(path.gene.id), "mmusculus", numeric_ns = "ENTREZGENE_ACC")

        tissue.path.loadings <- vector(mode = "list", length = length(transcript_loadings))
        names(tissue.path.loadings) <- tissue.names
        for(tx in 1:length(transcript_loadings)){
            gene.idx <- match(path.gene.table[,"target"], rownames(transcript_loadings[[tx]]))
            tissue.path.loadings[[tx]] <- transcript_loadings[[tx]][gene.idx]
        }

        #test <- aov_list(tissue.path.loadings)
        test <- lapply(tissue.path.loadings, function(x) t.test(x))
        pval_p <- sapply(test, function(x) x$p.value)
        
        vioplot(tissue.path.loadings, main = paste0(search.term, "\n", length(path.gene.id), " genes"), 
            ylim = c(all_min, all_max), col = tissue.cols)
        text(x = 1:4, y = all_max, labels = paste0("p = ", signif(pval_p, 2)))
        abline(h = 0)
    }
    dev.off()
}
```

### NES Heat Map

Plot a heatmap of the most extreme normalized enrichment scores together.

```{r enrich_heat, warning = FALSE, message = FALSE, error = FALSE, fig.width = 7, fig.height = 9}
#pdf(file.path(kegg.result.dir, "NES_heat.pdf"), width = 7, height = 50)
pheatmap(top.kegg.mat)
#dev.off()
```

### KEGG Pathway pictures {.tabset .tabset-fade .tabset-pills}

We can also plot loadings on individual pathway images. We use
the package pathview for this. Below we show NES overlaid onto
the Oxidative Phosphorylation pathway in each tissue.

In these pictures, the loadings are normalized to the min and
max of all transcript loadings across all tissues. So -1 corresponds 
to about -0.04 and 1 corresponds to about 0.04.

You can see below that OxPhos in adipose tissue is very negatively
loaded, meaning that greater expression is related to reduced
metabolic index (leaner mouse). OxPhos in the other tissues
is not loaded in any particular way, so it is not mediating
the genetic effects in the other tissues.

```{r oxphos, results = "asis", warning = FALSE, message = FALSE, error = FALSE}

search.term <- "Oxidative phosphorylation"

path.locale <- grep(search.term, all.kegg[[2]][,2], fixed = TRUE)
path.id <- all.kegg[[2]][path.locale,][1]
path.name <- all.kegg[[2]][path.locale,][2]

path.idx <- which(all.kegg[[1]][,1] == unlist(path.id))
path.gene.id <- all.kegg[[1]][path.idx,2]
path.gene.table <- gconvert(as.numeric(path.gene.id), "mmusculus", numeric_ns = "ENTREZGENE_ACC")

tissue.path.loadings <- vector(mode = "list", length = length(transcript_loadings))
names(tissue.path.loadings) <- tissue.names
for(tx in 1:length(transcript_loadings)){
    gene.idx <- match(path.gene.table[,"target"], rownames(transcript_loadings[[tx]]))
    gene.set <- transcript_loadings[[tx]][gene.idx]
    names(gene.set) <- path.gene.table[,"input"]
    gene.idx <- gene.idx[which(!is.na(gene.idx))]
    tissue.path.loadings[[tx]] <- gene.set
}

scale.factor <- max(sapply(transcript_loadings, function(x) max(abs(x), na.rm = TRUE)))

for(tx in 1:length(tissue.names)){
    cat("####", tissue.names[tx], "\n")
    pv.out <- pathview(gene.data = tissue.path.loadings[[tx]]/scale.factor, 
        pathway.id = path.id, species = "mmu", 
        out.suffix = paste(tissue.names[tx], gsub(" ", "_", search.term), sep = "_"),
        kegg.dir = here("Results", "Kegg"))

    picture.file <- paste0(path.id, ".", tissue.names[tx], "_", gsub(" ", "_", search.term), ".png")
    if(file.exists(here(picture.file))){
        system.call <- paste("mv", here(picture.file), file.path(kegg.result.dir, picture.file))
        system(system.call)
    }

    cat("![](",file.path(kegg.result.dir, picture.file), ")\n")

    cat("\n\n")
}
```


## GO Enrichments

We downloaded the GO annotation file (GAF) from MGI
\url{https://www.informatics.jax.org/downloads/reports/index.html}
We unzipped the file and saved mgi.gaf in Data/general

The column contents of a GAF are described here:
\url{https://geneontology.org/docs/go-annotation-file-gaf-format-2.2/}

```{r get_GO}
go.table <- read.delim(here("Data", "general", "mgi.gaf"), comment.char = "!", header = FALSE)
common.genes <- intersect(gene.table[,"external_gene_name"], go.table[,3])
gene.go.table <- go.table[which(go.table[,3] %in% common.genes),]

#remove NOT terms, since these have been shown NOT to be involved in the listed GO term
not.idx <- grep("NOT", gene.go.table[,4])
is.involved.idx <- setdiff(1:nrow(gene.go.table), not.idx)
is.go.table <- gene.go.table[is.involved.idx,]
```

```{r go_list}
min_gene_num  <- 80
max_gene_num  <- 3000

u_term <- unique(is.go.table[,5])
go.list <- lapply(u_term, function(x) unique(is.go.table[which(is.go.table[,5] == x),3]))
names(go.list) <- u_term
num.genes <- sapply(go.list, length)

good.idx <- intersect(which(num.genes >= min_gene_num), which(num.genes <= max_gene_num))
#length(good.idx)
go.list <- go.list[good.idx]

ensembl.go.list <- go.list
for(i in 1:length(go.list)){
    ensembl.id <- gene.table[match(go.list[[i]], gene.table[,"external_gene_name"]),"ensembl_gene_id"]
    ensembl.go.list[[i]]  <- ensembl.id
}

```


```{r go_names}
go.term <- names(ensembl.go.list)
term.ontology <- getGOOntology(go.term)
term.names <- goIdToTerm(go.term)
names(ensembl.go.list) <- term.names
```

```{r go_enrich}

go.enrich.nes.file <- here("Results", "GO", "GO_NES.RDS")

if(!file.exists(go.enrich.nes.file)){
    go.enrich.nes <- matrix(NA, ncol = length(tissue.names), nrow = length(go.list))
    colnames(go.enrich.nes) <- tissue.names
    rownames(go.enrich.nes) <- names(ensembl.go.list)

    for(tx in 1:length(tissue.names)){

        sorted.tx.loadings <- transcript_loadings[[tx]][order(transcript_loadings[[tx]], decreasing = TRUE),1]
        gsea.enrich <- fgsea::fgsea(ensembl.go.list, sorted.tx.loadings, scoreType = "std")
        norm.es <- as.numeric(as.matrix(gsea.enrich[,"NES"]))
        names(norm.es) <- gsea.enrich$pathway

        name.idx <- match(names(norm.es), rownames(go.enrich.nes))
        go.enrich.nes[name.idx,tx] <- norm.es
    }
    saveRDS(go.enrich.nes, go.enrich.nes.file)
}else{
    go.enrich.nes <- readRDS(go.enrich.nes.file)
}


vioplot(go.enrich.nes, col = tissue.cols)
abline(h = 0)
```

### Top GO Enrichments {.tabset .tabset-fade .tabset-pills}

Plot top and bottom enrichment scores.

```{r plot_go_enrich, fig.width = 9, fig.height = 7, results = "asis"}
#pdf(file.path(go.result.dir, "NES_by_tissue.pdf"), width = 9, height = 7)
num.top <- 10
extreme.go <- vector(mode = "list", length = length(tissue.names))
names(extreme.go) <- tissue.names

par(mar = c(4,4, 2, 4))
for(tx in 1:length(tissue.names)){
    cat("####", tissue.names[tx], "\n")
    enrich.order <- order(go.enrich.nes[,tx], decreasing = FALSE)
    sorted.enrich <- go.enrich.nes[enrich.order,tx]
    sorted.onto <- term.ontology[enrich.order]
    extreme.enrich  <- c(head(sorted.enrich, num.top), tail(sorted.enrich, num.top))
    extreme.onto <- c(head(sorted.onto, num.top), tail(sorted.onto, num.top))
    a <- barplot(extreme.enrich, 
        horiz = TRUE, las = 2, main = tissue.names[tx], 
        xlim = c(min(go.enrich.nes), max(go.enrich.nes)), names = NA)
    text(x = rep(-0.1, num.top), y = tail(a, num.top), 
        labels = paste0(tail(names(extreme.enrich), num.top), " (", tail(extreme.onto, num.top), ")"), adj = 1)
    text(x = rep(0.1, num.top), y = head(a, num.top), 
        labels = paste0(head(names(extreme.enrich), num.top), " (", head(extreme.onto, num.top), ")"), adj = 0)
    mtext("Normalized Enrichment Score", side = 1, line = 2.5)
 
    extreme.go[[tx]]  <- extreme.enrich
    cat("\n\n")
}
#dev.off()
```

### GO Term Loading Distributions

We plotted loading distributions for these top enriched terms
in GO_Term_Loadings.pdf

```{r all_terms}
top.go.enrich <- unique(unlist(lapply(extreme.go, names)))

pdf(file.path(go.result.dir, "GO_Term_Loadings.pdf"), width = 12, height = 12)
par(mfrow = c(4,4))
for(i in 1:length(top.go.enrich)){
    go.idx <- which(names(ensembl.go.list) == top.go.enrich[i])
    gene.vals <- lapply(transcript_loadings, function(x) x[match(ensembl.go.list[[go.idx]], rownames(x)),1])

    test <- lapply(gene.vals, function(x) t.test(x))
    pval_p <- sapply(test, function(x) x$p.value)
    
    vioplot(gene.vals, main = paste0(top.go.enrich[i], "\n", length(ensembl.go.list[[go.idx]])), 
        ylim = c(all_min, all_max), col = tissue.cols)
    text(x = 1:4, y = all_max, labels = paste0("p = ", signif(pval_p, 2)))
    abline(h = 0)
}
dev.off()
```

### GO term heat map

The following heat map shows the NES across the most 
extreme enriched terms

```{r go_enrich_heat, fig.width = 7, fig.height = 9}
top.idx <- match(top.go.enrich, rownames(go.enrich.nes))
pheatmap(go.enrich.nes[top.idx,])
```