---
title: "CC-RIX Metformin Effects"
author: Anna L Tyler
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  html_document:
    code_folding: hide
    collapsed: no
    toc: yes
    toc_float: yes
---

The purpose of this workflow is to explore the effects of 
metformin, diet, and sex, and the interactions between 
these things on gene expression in the CC-RIX.

This workflow is outside of the DO manuscript but depends 
on running the data parsing workflows for the CC-RIX.

Here we fit a linear model to each measured transcript 
that includes sex, diet, treatment, and the corresponding
imputed transcript (local genetic effect). This will tell 
us the contribution of each of these factors to the 
measured gene expression. 

The imputed gene expression, is our best guess at the 
genetic contribution to the level of each transcript.

Across the CC-RIX, the measured and imputed transcript
have some relationship. If this relationship is changed
by metformin, we can conclude that metformin is directly
changing the level of transcription. If we simply look
at differential expression in treated and untreated animals,
we do not know if metformin is directly changing expression 
or if the expression is changing due to some downstream
processes.


```{r get_args}
args <- commandArgs(trailingOnly = T)
tissue.name <- args[1]
cape_type <- args[2]
delete_previous <- as.logical(args[3])

if(is.na(tissue.name)){
    tissue.name = "adipose"
    cape_type = "query"
    #cape_type = "original"
    delete_previous <- FALSE
}
is.interactive <- FALSE
#is.interactive <- TRUE
```

This report shows results for `r tissue.name` tissue.

## Load data and code

```{r source_code}
library("here")
all.fun <- list.files(here("Code"), pattern = ".R", full.names = TRUE)
for(j in 1:length(all.fun)){source(all.fun[j])}
```


```{r load_libraries,  warning = FALSE, message = FALSE, error = FALSE}
#update.packages("vctrs", repos = "http://cran.us.r-project.org")

needed.packages <- c("pheatmap", "qtl2", "RColorBrewer", 
    "biomaRt", "stringr")
load_libraries(needed.packages)

#these are hard-coded so as not to interfere with the 
#HDM manuscript
load_latest_cape("~/Documents/git_repositories/cape")
#load query cape
query.fun <- list.files("~/Documents/git_repositories/query_cape", pattern = ".R", full.names = TRUE)
for(j in 1:length(query.fun)){source(query.fun[j])}
```

```{r set_dir}
data.dir <- here("Data", "CC-RIX") #where original data are stored
processed.data.dir <- here("Results", "CC-RIX", "Data") #where parsed data are stored
results.dir <- here("Results", "CC-RIX", "Gene_Expression_Effects") #where results from this workflow are stored
if(!file.exists(results.dir)){dir.create(results.dir)}
```


Get the expression for the specified tissue.


```{r read_data}
all.var <- ls()
data.loaded <- as.logical(length(which(all.var == "expr")))

if(!data.loaded){
    annot <- read.csv(file.path(data.dir, "final_sample_annotations.csv"))
    tissue.expr <- readRDS(file.path(processed.data.dir, "Expression.by.Tissue.RDS"))
    manifest <- as.matrix(read.csv(file.path(data.dir, "Original_Samples_Manifest.csv")))
    covar <- as.matrix(read.csv(file.path(processed.data.dir, "Covar.csv"), row.names = 1))
    bw <- as.matrix(read.csv(file.path(processed.data.dir, "Weight.Summarized.csv"), header = TRUE, row.names = 1))
    chem.mat <- read.csv(file.path(data.dir, "Blood.Chemistries.Summarized.csv"), row.names = 1)
    groups <- read.csv(file.path(processed.data.dir, "Groups.csv"), row.names = 1)
    group.def <- colnames(groups)

    gene.info.file <- here("Data", "general", "mouse_gene_info.txt")
    gene.info <- read.delim(gene.info.file)
    
}
```


```{r adjust, warning = FALSE, message = FALSE, error = FALSE}
tissue.locale <- which(names(tissue.expr) == tissue.name)

tissue.info <- unique(t(sapply(rownames(tissue.expr[[tissue.locale]]), 
  function(x) get_mouse_info(x, manifest, "Barcode.1"))))

orsam.id <- names(rownames(tissue.expr[[tissue.locale]]))
orsam.idx <- match(orsam.id, rownames(covar))

#adjust for age and RNASeq Batch
rnaseq.covar <- dummy_covar(covar[orsam.idx,c("Age", "RNAseq_Batch")])
adj.cc.expr <- adjust(tissue.expr[[tissue.locale]], rnaseq.covar)
saveRDS(adj.cc.expr, file.path(processed.data.dir, paste0("Adjusted.Expression.", tissue.name, ".RDS")))

mouse.info <- unique(t(sapply(rownames(covar), function(x) get_mouse_info(x, manifest, "User.Inputs"))))
rownames(mouse.info) <- mouse.info[,1]

#adjust phenotypes for age
pheno.covar <- dummy_covar(mouse.info[,"Timepoint",drop=FALSE])
num.chem <- apply(as.matrix(chem.mat[,5:ncol(chem.mat)]), 2, as.numeric)
rownames(num.chem) <- rownames(chem.mat)
adj.chem <- adjust(num.chem, pheno.covar)
saveRDS(adj.chem, file.path(processed.data.dir, "Adjusted.Chem.RDS"))
adj.bw <- adjust(bw, pheno.covar)
saveRDS(adj.bw, file.path(processed.data.dir, "Adjusted.BW.RDS"))
```


## Metformin and Traits

Does metformin affect the traits overall?
Nope.

```{r met_traits}
log.chem <- log10(round(adj.chem+1))
zero.idx <- which(log.chem == 0, arr.ind = TRUE)
#adj.chem[zero.idx]
log.chem[which(log.chem == 0)] <- NA

scaled.chem <- apply(log.chem, 2, scale)
boxplot(scaled.chem)
chem.cov <- cov(scaled.chem, use = "pairwise.complete.obs")
pheatmap(chem.cov)
chem.decomp <- plot.decomp(chem.cov, label.points = TRUE)
ind.cov <- cov(t(scaled.chem), use = "pairwise.complete.obs")
plot.decomp(ind.cov)

chem.idx <- match(rownames(adj.chem), tissue.info[,1])
chem.info <- tissue.info[chem.idx,]

chem.df <- data.frame("Sex" = as.factor(chem.info[,"Sex"]), 
        "Diet" = as.factor(chem.info[,"Diet"]),
        "Treatment" = as.factor(chem.info[,"Treatment"]),
        "Strain" = as.numeric(as.factor(chem.info[,"User.Defined.Strain"])))

outliers1 <- which(chem.decomp$u[,1] > 0.2)
outliers2 <- which(chem.decomp$u[,2] > 0.2)

chem.info[outliers1,]
adj.chem[outliers1,]

chem.info[outliers2,]
adj.chem[outliers2,]

outlier.idx <- rep(0, nrow(chem.info))
outlier.idx[outliers1] <- 1
outlier.idx[outliers2] <- 2
outlier.df <- data.frame("outliers" = as.factor(outlier.idx))
rownames(outlier.df) <- rownames(log.chem)
chem.cor <- cor(t(scaled.chem), use = "pairwise.complete.obs")
rownames(chem.cor) <- colnames(chem.cor) <- rownames(log.chem)
pheatmap(chem.cor, annotation_row = outlier.df, show_rownames = FALSE, show_colnames = FALSE)

test <- cov(t(scaled.chem), use = "pairwise.complete.obs")
plot.decomp(test)

plot.decomp(chem.cor, cols = chem.df[,3])
```

## Group Expression

Group the measured expression based on the CC-RIX crosses.

```{r group_expr}
cross.id <- tissue.info[,"User.Defined.Strain"]
u_cross <- unique(cross.id)
cross.idx <- lapply(u_cross, function(x) which(cross.id == x))
cross_expr <- lapply(cross.idx, function(x) adj.cc.expr[x,])
names(cross_expr) <- u_cross
mean.cross.expr <- t(sapply(cross_expr, colMeans))
```

## Expression Decomposition

The following plots show the decompmosition of the gene expression
matrix. Each point represents one individual. Its strain ID is 
indicated by a number. Points are colored by Sex, Diet, or Treatment.

```{r expr_decomp, fig.width = 8, fig.height = 8}
expr.decomp <- plot.decomp(adj.cc.expr, plot.results = FALSE)
xlab = paste0("PC1 (", signif(expr.decomp$var.exp[1], 2)*100, "%)")
ylab = paste0("PC2 (", signif(expr.decomp$var.exp[2], 2)*100, "%)")

if(is.interactive){quartz(width = 12, height = 12)}
par(mfrow = c(2,2))
plot(expr.decomp$u, xlab = xlab, ylab = ylab, main = "Colored by Diet", 
    cex = 1.5, col = "darkgray")
text(expr.decomp$u[,1], expr.decomp$u[,2], 
    labels = as.numeric(as.factor(tissue.info[,"User.Defined.Strain"])),
    col = as.numeric(as.factor(tissue.info[,"Diet"])), cex = 0.5)

plot(expr.decomp$u, xlab = xlab, ylab = ylab, main = "Colored by Sex", 
    cex = 1.5, col = "darkgray")
text(expr.decomp$u[,1], expr.decomp$u[,2], 
    labels = as.numeric(as.factor(tissue.info[,"User.Defined.Strain"])),
    col = as.numeric(as.factor(tissue.info[,"Sex"])), cex = 0.5)

plot(expr.decomp$u, xlab = xlab, ylab = ylab, main = "Colored by Treatment", 
    cex = 1.5, col = "darkgray")
text(expr.decomp$u[,1], expr.decomp$u[,2], 
    labels = as.numeric(as.factor(tissue.info[,"User.Defined.Strain"])),
    col = as.numeric(as.factor(tissue.info[,"Treatment"])), cex = 0.5)

```

For a given group with the same cofactors, does expression
cluster by strain? Sort of. In the plots below, each strain
is indicated by a number 1 through 20. Individuals from a
given strain do tend to appear in the same region of the 
plot, however, there isn't the same tight clustering of 
strains that we see in the inbred founders. I think this
is because CAST and PWK are so so so different from the
domesticus strains that they blow up the scale on which
we are plotting expression differences. Strains appear
to cluster very tightly because there is so much distance
between them. In the CC-RIX, all animals have a mix of 
CAST, PWK, and domesticus strains, so there are no individual
mice that are very distant from the others. Individuals do 
appear near each other, but because we are zoomed in to the
level of noise, the strains don't appear to cluster as 
tightly as they would if there were wildy divergent strains.
I bet if you put BxD mice in the same plot, they would cluster
well away from the CC-RIX in a way that distinctly separated
the two populations. As it is, though, all CC-RIX are pretty
close to each other in transcription space.

```{r group_pc, fig.width = 8, fig.height = 8}
u_groups <- unique(tissue.info[,c("Sex", "Diet", "Treatment")])
group.idx <- apply(u_groups, 1, function(x) Reduce("intersect", 
    list(which(tissue.info[,"Sex"] == x[1]), which(tissue.info[,"Diet"] == x[2]), 
    which(tissue.info[,"Treatment"] == x[3]))))
group.expr <- lapply(group.idx, function(x) adj.cc.expr[x,])

#pdf("~/Desktop/Group_Decomp.pdf", width = 16, height = 8)
par(mfrow = c(2,2))
for(i in 1:length(group.expr)){
    group.decomp <- plot.decomp(group.expr[[i]], plot.results = FALSE)
    xlab = paste0("PC1 (", signif(group.decomp$var.exp[1], 2)*100, "%)")
    ylab = paste0("PC2 (", signif(group.decomp$var.exp[2], 2)*100, "%)")
    plot(group.decomp$u, cex = 2.7, main = paste(u_groups[i,], collapse = ", "),
        xlab = xlab, ylab = ylab)
    text(group.decomp$u[,1], group.decomp$u[,2], 
        labels = as.numeric(as.factor(tissue.info[group.idx[[i]],"User.Defined.Strain"])))
}
#dev.off()
```

For each measured transcript, we fit a model that indludes Sex, Diet, 
Treatment, and strain, as well as the interaction between strain and 
each other cofactor.

```{r model_with_strain}

matched.covar <- get.xz(adj.cc.expr, mouse.info)$Z

model.coef.file <- file.path(results.dir, paste0("model_strain_coef_", tissue.name, ".RDS"))
model.p.file <- file.path(results.dir, paste0("model_strain_p_", tissue.name, ".RDS"))

if(!file.exists(model.coef.file) || !file.exists(model.p.file) || delete_previous){
    all.coef <- all.p <- matrix(NA, nrow = ncol(adj.cc.expr), ncol = 8)
    rownames(all.coef) <- rownames(all.p) <- colnames(adj.cc.expr)

    for(i in 1:ncol(adj.cc.expr)){
        if(is.interactive){report.progress(i, ncol(adj.cc.expr))}

        df <- data.frame(cbind(adj.cc.expr[,i], as.factor(matched.covar[,"Sex"]), 
            as.factor(matched.covar[,"Diet"]), as.factor(matched.covar[,"Treatment"]), 
            as.factor(matched.covar[,"User.Defined.Strain"])))
        colnames(df) <- c("measured_expression", "Sex", "Diet", "Treatment", "Strain")

        model <- lm(measured_expression~Sex+Diet+Treatment+Strain+Sex:Strain+Diet:Strain+Treatment:Strain, data = df)
        #summary(model)

        model.sum <- summary(model)
        all.coef[i,] <- model.sum$coefficients[,"Estimate"]
        all.p[i,] <- model.sum$coefficients[,"Pr(>|t|)"]
    }
    colnames(all.coef) <- colnames(all.p) <- rownames(model.sum$coefficients)
    saveRDS(all.coef, model.coef.file)
    saveRDS(all.p, model.p.file)
}else{
all.coef <- readRDS(model.coef.file)
all.p <- readRDS(model.p.file)
}
```

The following box plot shows the distribution of the coefficients
for each factor. This plot shows that sex and diet have the largest
effects on transcription, as seen above in the expression matrix 
decompositions. Interestingly, strain has an even smaller effect
than metformin treatment, which already has a pretty minimal effect.

```{r coef_dist}
par(mar = c(8, 4, 2, 2))
boxplot(abs(all.coef[,2:ncol(all.coef)]), las = 2)
```

The following plots show the p value distributions for each set of 
factors. These are all inflated, except for Treatment, and the 
Treatment by strain effects. I think this is because all the rest
of the factors have widespread effects on transcription. So the 
inflation is just an indication of the widespread effects, not 
an indication that we've done something wrong. We do expect each
of these factors to have widespread effects.

```{r qqunif, fig.width = 10, fig.height = 6}
par(mfrow = c(2,4))
for(i in 2:ncol(all.p)){
    qqunif.plot(all.p[,i], plot.label = colnames(all.p)[i])
}
```

## Effects summary

The following plots show the beta coefficients plotted against
the -log(fdr) for all independent variables in the model. Each
dot is a transcript

```{r effect_summary, fig.width = 10, fig.height = 6}
fdr.level = 0.1
fdr <- apply(all.p, 2, function(x) p.adjust(x, "fdr"))
layout.mat <- matrix(c(rep(1:3, each = 4), rep(4:7, each = 3)), nrow = 2, byrow = TRUE)
layout(layout.mat)
for(i in 2:ncol(all.coef)){
    plot(all.coef[,i], -log(fdr[,i]), main = colnames(all.coef)[i], 
    xlab = "Beta coefficient", ylab = "-log(p value)")
    abline(h = -log(fdr.level))
    #abline(h = -log(0.05/nrow(all.coef)))
}
```

## Number of Significant Effects

The following plot shows the number of significant effects per
category.

```{r num_sig}
sig.transcripts <- apply(fdr, 2, function(x) rownames(all.coef)[which(x <= fdr.level)])
sig.up <- lapply(1:ncol(fdr), function(x) rownames(all.coef)[intersect(which(fdr[,x] <= fdr.level), which(all.coef[,x] > 0))])
sig.down <- lapply(1:ncol(fdr), function(x) rownames(all.coef)[intersect(which(fdr[,x] <= fdr.level), which(all.coef[,x] < 0))])
names(sig.up) <- names(sig.down) <- colnames(fdr)

num.sig <- sapply(sig.transcripts, length)
num.up <- sapply(sig.up, length)
num.down <- sapply(sig.down, length)
par(mar = c(4, 14, 2, 2))
barplot_with_num(num.sig, las = 2, horiz = TRUE, text.gap = 0.15, text.shift = 0.07)
```

## Enrichment of Genes Differentially Expressed by Treatment {.tabset .tabset-fade .tabset-pills}

The following plot shows functional enrichment for transcripts that
were significantly up or downregulated by metformin.

```{r up_down_treat, fig.width = 8, fig.height = 7, results = "asis", eval = FALSE}
library(gprofiler2)
treat.up.enrich <- gost(sig.up$Treatment, organism = "mmusculus")
treat.down.enrich <- gost(sig.down$Treatment, organism = "mmusculus")

cat("### Up-regulated by Metformin\n")
plot.enrichment(treat.up.enrich, num.terms = 20, , max.term.size = 4000,
    plot.label = "Upregulated by Metformin", order.by = "p_value")
cat("\n\n")

cat("### Down-regulated by Metformin\n")
plot.enrichment(treat.down.enrich, num.terms = 20, , max.term.size = 4000,
    plot.label = "Downregulated by Metformin", order.by = "p_value")
cat("\n\n")

cat("### Interacting with Imputed Transcript\n")
enrich <- gost(sig.transcripts[[8]], organism = "mmusculus")
plot.enrichment(enrich, num.terms = 20, order.by = "p_value")
```

## Examples {.tabset .tabset-fade .tabset-pills}

The following plots show a more detailed picture of some of
the imputed transcripts whose relationship to the measured
transcript was changed by metformin.

```{r example}
plot_int <- function(x.factor, int.factor, response, xlab = "", ylab = "", 
    trace.label = "", main = ""){

    df <- data.frame(cbind(response, x.factor, int.factor))
    colnames(df) <- c("response", "x", "y")
    model <- lm(response~x*y, data = df)
    #summary(model)
	p <- signif(summary(model)$coefficients[4,4], 2)

    group.x <- levels(x.factor)
    group.int <- levels(int.factor)
    
    comp.ids <- rbind(cbind(rep(group.x, each = length(group.int)), group.int))
    group.names <- apply(comp.ids, 1, function(x) paste(x, collapse = "_"))
    comp.idx <- apply(comp.ids, 1, function(x) intersect(which(x.factor == x[1]), which(int.factor == x[2])))

    group.vals <- lapply(comp.idx, function(x) response[x])
    group.means <- sapply(group.vals, mean)
    group.se  <- sapply(group.vals, function(x) sd(x)/sqrt(length(x)))
    names(group.vals) <- names(group.means) <- names(group.se) <- group.names

    ylim <- c(min(group.means-group.se), max(group.means+group.se))

    error.x <- rep(1:length(group.x), each = length(group.int))

    interaction.plot(x.factor, int.factor, response, 
        main = paste0(main, " (p = ", signif(p, 2), ")"), ylim = ylim,
        xlab = xlab, ylab = ylab, trace.label = trace.label)
    segments(error.x, group.means-group.se, error.x, group.means+group.se)
    
}


plot_example <- function(transcript.id){
    
    gene.name <- gene.info[which(gene.info[,1] == transcript.id),"external_gene_name"]
    sex.diet.covar <- dummy_covar(matched.covar[,c("Sex", "Diet")])

    par(mfrow = c(1,3), mar = c(4,4,6,4))

    model <- lm(cc.measured[,transcript.id]~as.factor(matched.covar[,"Sex"]))
    f <- summary(model)$fstatistic
	p <- signif(pf(f[1],f[2],f[3],lower.tail=F), 2)
    plot(cc.measured[,transcript.id]~as.factor(matched.covar[,"Sex"]),
    xlab = "Sex", ylab = "Measured Expression", main = paste("Effect of Sex\np =", p))

    model <- lm(cc.measured[,transcript.id]~as.factor(matched.covar[,"Diet"]))
    f <- summary(model)$fstatistic
	p <- signif(pf(f[1],f[2],f[3],lower.tail=F), 2)
    plot(cc.measured[,transcript.id]~as.factor(matched.covar[,"Diet"]),
    xlab = "Diet", ylab = "Measured Expression", main = paste("Effect of Diet\np =", p))

    model <- lm(cc.measured[,transcript.id]~as.factor(matched.covar[,"Treatment"]))
    f <- summary(model)$fstatistic
	p <- signif(pf(f[1],f[2],f[3],lower.tail=F), 2)
    plot(cc.measured[,transcript.id]~as.factor(matched.covar[,"Treatment"]),
    xlab = "Treatment", ylab = "Measured Expression", main = paste("Effect of Treatment\np =", p))

    int.group <- unique(matched.covar[,"Treatment"])
    treat1.locale <- which(matched.covar[,"Treatment"] == int.group[1])
    treat2.locale <- which(matched.covar[,"Treatment"] == int.group[2])
    ylim <- c(min(cc.measured[,transcript.id]), max(cc.measured[treat1.locale,transcript.id]))

    mar = c(4,4,2,4)

    #plot.with.model(cc.imputed[treat1.locale,transcript.id], 
    #cc.measured[treat1.locale,transcript.id],
    #xlab = "Imputed Expression", ylab = "Measured Expression", 
    #main = int.group[1], ylim = ylim, report = "cor.test")
    
    #plot.with.model(cc.imputed[treat2.locale,transcript.id], 
    #cc.measured[treat2.locale,transcript.id],
    #xlab = "Imputed Expression", ylab = "Measured Expression", 
    #main = int.group[2], ylim = ylim, report = "cor.test")

    #adjust the measured expression for sex and diet
    adj.measured <- adjust(cc.measured[,transcript.id,drop=FALSE], sex.diet.covar)

    #bin the imputed expression into three bins
    #binned.imputed <- round(bin.vector(cc.imputed[,transcript.id], 
    #    segment.region(min(cc.imputed[,transcript.id]), max(cc.imputed[,transcript.id]), 3)), 2)

    #plot_int(x.factor = as.factor(binned.imputed), 
    #    int.factor = as.factor(matched.covar[,"Treatment"]), 
    #    response = adj.measured, xlab = "Imputed Expression", 
    #    ylab = "Measured Expression", trace.label = "Treatment",
    #    main = "Interaction") 
    
    mtext(gene.name, side = 3, outer = TRUE, line = -1.5, font = 2)

}
```


```{r example_plots, results = "asis"}
cc.measured <- adj.cc.expr
example.id <- rownames(fdr)[which(fdr[,8] <= 0.2)]

#example.idx <- sample(1:nrow(fdr), 5); names(example.idx) <- rownames(fdr)[example.idx]
num.examples <- min(length(example.id), 18)

for(i in 1:num.examples){
    gene.name <- gene.info[which(gene.info[,1] == example.id[i]), "external_gene_name"]
    cat("###", gene.name, "\n")
    if(is.interactive){quartz(width = 8, height = 6)}
    plot_example(example.id[i])
    cat("\n\n")
}


#gene.name = "E2f7"
#gene.name = "BC067074"
#gene.id <- gene.info[which(gene.info[,"external_gene_name"] == gene.name),1]
#gene.id <- "ENSMUSG00000019787"
#plot_example(gene.id)
```
