On strain separation in transcriptome space

For a given group with the same cofactors, does expression
cluster by strain? Sort of. In the plots below, each strain
is indicated by a number 1 through 20. Individuals from a
given strain do tend to appear in the same region of the 
plot, however, there isn't the same tight clustering of 
strains that we see in the inbred founders. I think this
is because CAST and PWK are so so so different from the
domesticus strains that they blow up the scale on which
we are plotting expression differences. Strains appear
to cluster very tightly because there is so much distance
between them. In the CC-RIX, all animals have a mix of 
CAST, PWK, and domesticus strains, so there are no individual
mice that are very distant from the others. Individuals do 
appear near each other, but because we are zoomed in to the
level of noise, the strains don't appear to cluster as 
tightly as they would if there were wildy divergent strains.
I bet if you put BxD mice in the same plot, they would cluster
well away from the CC-RIX in a way that distinctly separated
the two populations. As it is, though, all CC-RIX are pretty
close to each other in transcription space.


plotting imputed and measured interactions

```{r}
#plot 7 measured by imputed in no treatment group
    par(mar = c(4,4,4,2))
    plot.with.model(cc.imputed[treat1.locale,transcript.id], 
    cc.measured[treat1.locale,transcript.id],
    xlab = "Imputed Expression", ylab = "Measured Expression", 
    main = int.group[1], ylim = ylim, report = "cor.test")
    
    #plot 8 measured by imputed in treatment group
    plot.with.model(cc.imputed[treat2.locale,transcript.id], 
    cc.measured[treat2.locale,transcript.id],
    xlab = "Imputed Expression", ylab = "Measured Expression", 
    main = int.group[2], ylim = ylim, report = "cor.test")

    #adjust the measured expression for sex and diet
    adj.measured <- adjust(cc.measured[,transcript.id,drop=FALSE], sex.diet.covar)

    #bin the imputed expression into three bins
    binned.imputed <- round(bin.vector(cc.imputed[,transcript.id], 
        segment.region(min(cc.imputed[,transcript.id]), max(cc.imputed[,transcript.id]), 3)), 2)

    #plot 9 metformin expression interaction
    plot_int(x.factor = as.factor(binned.imputed), 
        int.factor = as.factor(matched.covar[,"Treatment"]), 
        response = adj.measured, xlab = "Imputed Expression", 
        ylab = "Measured Expression", trace.label = "Treatment",
        main = "Interaction") 
```

From when we were fitting the imputed transcript. I can't get my head
around why we did it, so I am taking it out for now

## Model Fitting

For each measured transcript we fit a model including Sex, Diet, Treatment,
Timepoint, Strain, and all the possible interactions.


```{r fit_models}
common.transcripts <- intersect(colnames(adj.cc.expr), colnames(ind.auto.expr)[has.vals])
common.cc.idx <- match(common.transcripts, colnames(adj.cc.expr))
common.imp.idx <- match(common.transcripts, colnames(ind.auto.expr))

cc.measured <- adj.cc.expr[,common.cc.idx]
cc.imputed <- ind.auto.expr[,common.imp.idx]
rownames(cc.imputed) <- rownames(cc.measured) #make sure both have the same rownames

matched.covar <- get.xz(cc.measured, mouse.info)$Z

model.coef.file <- file.path(results.dir, paste0("model_coef_", tissue.name, ".RDS"))
model.p.file <- file.path(results.dir, paste0("model_p_", tissue.name, ".RDS"))

if(!file.exists(model.coef.file) || !file.exists(model.p.file) || delete_previous){
    all.coef <- all.p <- matrix(NA, nrow = ncol(cc.measured), ncol = 10)
    rownames(all.coef) <- rownames(all.p) <- common.transcripts

    for(i in 1:length(common.transcripts)){
        if(is.interactive){report.progress(i, length(common.transcripts))}

        df <- data.frame(cbind(cc.measured[,i], as.factor(matched.covar[,"Sex"]), 
            as.factor(matched.covar[,"Diet"]), as.factor(matched.covar[,"Timepoint"]),
            as.factor(matched.covar[,"Treatment"]), 
            as.factor(matched.covar[,"User.Defined.Strain"]),
            cc.imputed[,common.transcripts[i]]))
        colnames(df) <- c("measured_expression", "Sex", "Diet", 
            "Timepoint", "Treatment", "Strain", "imputed_expression")

        model <- lm(measured_expression~Sex+Diet+Timepoint+Treatment+imputed_expression+Sex:imputed_expression+Diet:imputed_expression+Timepoint:imputed_expression+Treatment:imputed_expression, data = df)
        
        #summary(model)

        model.sum <- summary(model)
        all.coef[i,] <- model.sum$coefficients[,"Estimate"]
        all.p[i,] <- model.sum$coefficients[,"Pr(>|t|)"]
    }
    colnames(all.coef) <- colnames(all.p) <- rownames(model.sum$coefficients)
    saveRDS(all.coef, model.coef.file)
    saveRDS(all.p, model.p.file)
}else{
all.coef <- readRDS(model.coef.file)
all.p <- readRDS(model.p.file)
}
```
